---
title: "multical"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multical}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# `multical`: Multilevel calibration weighting for surveys

## Installation
You can install `multical` from github using `devtools`.

```{r install, results="hide", message=F, eval=F}
## Install devtools if not already installed
install.packages("devtools", repos='http://cran.us.r-project.org')
## Install multical from github
devtools::install_github("ebenmichael/multical")
```


## Data

`multical` takes two separate data frames as inputs:

- **`sample_data`**: individual-level data for survey respondents. One row per respondent.
- **`pop_data`**: data describing the target population. This can be individual-level (one row per population member) or pre-aggregated to the cell level (one row per unique combination of covariate values, with a column giving the count per cell).

We'll use two contrived examples shipped with the package: `data_individual` has individual-level information on covariates, response, and outcome; `data_cell` has cell-level information.

```{r setup}
library(multical)
library(dplyr)
data(data_individual)
data(data_cell)
```

`data_individual` records 4 covariates, whether each individual responded (`response`) to the survey, whether they are in the target population (`intarget`), and their answer to a binary question `y`.

```{r indiv}
head(data_individual)
```

`data_cell` records each unique combination of the 4 covariates, the number of respondents in that cell (`sample_count`), the number of individuals in the target population in that cell (`target_count`), and the mean outcome in the cell.

```{r cell}
head(data_cell)
```

For the examples below we'll split `data_individual` into respondents (our sample) and the full population.

```{r split}
sample_ind <- data_individual %>% filter(response == 1)
pop_ind    <- data_individual   # all rows; each counts as 1 population member
```


## Getting multilevel calibration weights

The main function is `multical`. Its first three arguments are:

1. `formula`: a **right-hand-side only** formula (`~ covariates`) specifying which variables to calibrate on
2. `sample_data`: the respondent data frame
3. `pop_data`: the population data frame (individual or cell level)

An optional `target_count` argument names a column in `pop_data` giving the count or weight per row. When omitted, each row of `pop_data` is treated as one population member.

### Raking on first-order margins (individual-level population)

Setting `order = 1` calibrates only on the main-effect margins of each covariate.

```{r multical_indiv_rake}
out <- multical(~ X1 + X2 + X3 + X4, sample_ind, pop_ind, order = 1)
```

The output has one row per respondent **plus** one row per covariate cell that exists in the population but has no respondents (`sample_count = 0`, `weight = 0`). The extra rows are needed for correct balance calculations; filter them out before estimation.

Any auxiliary outcome columns (e.g. `y`) are not carried through by `multical`. Because the respondent rows in the output are in the same order as the rows of `sample_data`, you can retrieve them with `bind_cols` after filtering to respondents.

We can inspect marginal balance using `get_balance`:

```{r multical_indiv_rake_imbal}
get_balance(out, 1)
```

To estimate the population average, filter to respondents and take the weighted mean:

```{r estimate_indiv_rake}
out %>%
  filter(sample_count != 0) %>%
  bind_cols(sample_ind %>% select(y)) %>%
  summarise(estimate = sum(weight * y) / sum(weight))
```

### Higher-order calibration with a fixed lambda

Setting `order > 1` includes interaction terms up to that order. The parameter `lambda` controls the trade-off between exact post-stratification (small `lambda`) and lower variance (large `lambda`).

```{r multical_indiv}
out <- multical(~ X1 + X2 + X3 + X4, sample_ind, pop_ind, order = 4, lambda = 1)

rbind(head(get_balance(out, 4)), tail(get_balance(out, 4)))
```

### Selecting lambda automatically

By default (no `lambda` supplied), `multical` solves over a grid of values from `lambda_max` down to `lambda_max * lambda_min_ratio`. Use `get_balance_v_sample_size` to trace the balance-effective-sample-size trade-off and choose an appropriate value.

```{r multical_indiv_lambda}
out <- multical(~ X1 + X2 + X3 + X4, sample_ind, pop_ind)

imbal <- get_balance_v_sample_size(out, 4)
plot(imbal$n_eff, imbal$imbalance,
     xlab = "Effective sample size", ylab = "Normalized imbalance")
```

For large datasets with many covariates, start with a low `order` and increase as needed — the number of interaction terms grows quickly.

### Using pre-aggregated cell-level population data

If the population is available as a pre-aggregated cell table (e.g. from a census), pass it as `pop_data` and name the count column via `target_count`. The sample is still individual-level.

```{r multical_cell}
out <- multical(~ X1 + X2 + X3 + X4, sample_ind, data_cell,
                target_count = target_count,
                order = 4, lambda = 1)
```

Estimation works the same way — filter to respondents and take the weighted mean:

```{r multical_cell_estimate}
out %>%
  filter(sample_count != 0) %>%
  bind_cols(sample_ind %>% select(y)) %>%
  summarise(estimate = sum(weight * y) / sum(weight))
```